Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index

### 2016-05-18 swsc.tex
Updates from ShareLaTeX

The magnetosphere's dynamics and its associated solar wind driver form a complex nonlinear coupled dynamical system. It is therefore not only instructive but also greatly simplifying to use representative indices to quantify the state of geomagnetic activity.

Geomagnetic indices come in various forms, they may take continuous or discrete values. Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomena. Some popular indices are the $K_p$, $Dst$ and the $AE$ index.

\begin{comment}
\begin{enumerate}
    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices, for a given value, for each of the past days. The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.

    \item AE: The Auroral Electrojet Index, AE, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval.

    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents. 
\end{enumerate}
\end{comment}

%Talk about Burton and friends
We focus on prediction of the hourly $Dst$ index which is an important problem because $Dst$ is a straightforward indicator of geomagnetic storms. More specifically we focus on the \emph{one step ahead} (OSA) prediction of $Dst$ because it is the first step towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts so far, some of the earliest are by \citet{JGR:JGR10260} who outlined a prediction framework that has been used as a starting point for many $Dst$ models. \citet{JGR:JGR10260} model $Dst(t)$ a function of time using an \emph{Ordinary Differential Equation} (ODE) which expressed it's rate of change as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$.

The \citet{JGR:JGR10260} model has proven to be very influential in the Space Weather research effort and many subsequent works like \citet{Wang:Dst}, \citet{JGRA:JGRA14856} modified the proposed ODE by proposing expressions for the injection term $Q(t)$. More recently \citet{Ballatore2014} have tried to generate empirical estimates for the injection and decay terms in Burton's equation.

%Talk about NARMAX Dst
Among the other major empirical models used to predict $Dst$ has been the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations or monomials to include in the refined model using the \emph{error reduction ratio} (ERR) criterion. The parameters of the NARMAX OLS-ERR model are calculated using the standard \emph{ordinary least squares} (OLS) method hence these methods are also denoted by the acronym NARMAX OLS-ERR. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

%Talk about neural networks
\emph{Neural Networks} have been a popular choice for building predictive models and researchers employed the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $Dst$, which used the solar wind, \emph{interplanetary magnetic field} (IMF) and historical $Dst$ data. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} (\citet{boyle1997empirical}), in \citet{SWE:SWE286} the same architecture is adapted for prediction of $Dst$, which is popularly known as the Rice $Dst$ model. \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.

%Talk about comparison papers, number of storms covered and Persistence
In light of the extensive list of modeling techniques employed for prediction of geomagnetic indices and in particular $Dst$, model comparison and evaluation becomes a crucial direction in advancing the research domain. \citet{Rastatter2013} compare several physics based (convection, kinetic and magneto hydrodynamic) and empirical prediction models such as Narmax, Rice $Dst$ on 4 storm events which occurred between 2001 and 2006. \citet{Amata2008} compares the EDDA and the Lund $Dst$ models over the 2003-2005 period. But the most extensive model comparison in terms of storm events was conducted in \citet{Ji2012} which compared seven $Dst$ models (see table \ref{table:DstModels}) on a list of 63 geomagnetic storm events of varying intensities which occurred between 1998 and 2006.

>>>>

The magnetosphere's dynamics and its associated solar wind driver form a complex dynamical system. It is therefore instructive and greatly simplifying to use representative indices to quantify the state of geomagnetic activity.

Geomagnetic indices come in various forms, they may take continuous or discrete values, indices may be calculated at varying time frequencies. Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth. Each geomagnetic index is a proxy for a particular kind of phenomena. Some popular indices are the $K_p$, $Dst$ and the $AE$ index.

\begin{enumerate}
    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices, for a given value, for each of the past days. The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.

    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval. It is a continuous index which is calculated every hour.

    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents. 
\end{enumerate}

%Talk about Burton and friends
We focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms. More specifically, we focus on the \emph{one step ahead} (OSA) prediction of $Dst$ because it is the first piece towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature. One the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$.

The \citet{JGR:JGR10260} model has proven to be very influential particularly due to its simplicity. Many subsequent works have modified the proposed ODE by proposing alternative expressions for the injection term $Q(t)$ [see \citet{Wang:Dst}, \citet{JGRA:JGRA14856}]. More recently \citet{Ballatore2014} have tried to generate empirical estimates for the injection and decay terms in Burton's equation.

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated using the standard \emph{ordinary least squares} (OLS) method to minimize an objective function. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

%Talk about neural networks
Yet another family of forecasting methods is based on \emph{Artificial Neural Networks} (ANN) that have been a popular choice for building predictive models. Researchers have employed both the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures. \citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $Dst$, which used the solar wind velocity, \emph{interplanetary magnetic field} (IMF) and historical $Dst$ data as inputs. \citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} [see \citet{boyle1997empirical}]. The same architecture is adapted for prediction of $Dst$ in \citet{SWE:SWE286}, which is popularly known as the Rice $Dst$ model. \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.

%Talk about comparison papers, number of storms covered and Persistence
In light of the extensive list of modeling techniques employed for prediction of the $Dst$ index, model comparison and evaluation becomes a crucial direction in advancing the research domain. \citet{Rastatter2013} compare several physics based (convection, kinetic and magneto hydrodynamic) and empirical prediction models such as NARMAX, Rice $Dst$ on 4 storm events which occurred between 2001 and 2006. \citet{Amata2008} compare the EDDA and the Lund $Dst$ models over the 2003-2005 period. The most extensive model comparison in terms of storm events was probably conducted in \citet{Ji2012} which compared six $Dst$ models (see table \ref{table:DstModels}) on a list of 63 geomagnetic storm events of varying intensities which occurred between 1998 and 2006. In the comparison done in \citet{Ji2012}, the model proposed in \citet{JGRA:JGRA16300} gives the best performance on the test set considered.
